---
title: "p8105_hw2_yc4959"
author: "Yimeng Cao"
date: "2026-09-30"
output: github_document
---

This is p8105 homework 2

```{r}
library(tidyverse)
```

## Problem 1

Read and clean data, select retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable.

So I need to remove the exit_only column.
```{r}
subway_data = 
  read_csv("./NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>% 
  janitor::clean_names() %>% 
  select(line:vending, - exit_only, ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

This subway dataset contains line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance.
I cleaned the variable names, selected the needed variables, removed the exit_only variable, and converted entry from character to logical.

The cleaned dataset contains `r nrow(subway_data)` rows and `r ncol(subway_data)` columns.

I don't think this is “tidy data” because the data for routes 1 through 11 is spread across different columns.


How many distinct stations are there? 
```{r}
subway_data |> 
  distinct(line, station_name) |> 
  nrow()
```
465 distinct stations are there.

How many stations are ADA compliant?
```{r}
subway_data |> 
  distinct(line, station_name, ada) |> 
  count(ada)

```
84 stations are ADA compliant.

What proportion of station entrances / exits without vending allow entrance?

Use no vending entry divided by total no vending entry
```{r}
no_vending_entry = 
  subway_data |> 
   filter(vending == "NO", entry == TRUE) |> 
  nrow()

no_vending_total =
   subway_data |> 
  filter(vending == "NO") |> 
  nrow()

proportion = no_vending_entry / no_vending_total 
proportion

```
The proportion of station entrances / exits without vending allow entrance is 37.7%.



Reformat data so that route number and route name are distinct variables.

When reshaping the data, I found that the route variables had different data types. Route1 to Route7 were stored as character variables, while Route8 to Route11 were stored as numeric variables.
So i use mutate to change the data type.
```{r}
subway_data_reformatted =
  subway_data |> 
  mutate(
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
  ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  )
```

How many distinct stations serve the A train?
```{r}
distinct_A_stations =
  subway_data_reformatted |> 
    filter(route_name == "A") |>
   distinct(line, station_name) |> 
  nrow()
distinct_A_stations
```
60 stations serve the A train.

How many distinct stations serve the A train?
```{r}
distinct_A_stations =
  subway_data_reformatted |> 
    filter(route_name == "A") |>
   distinct(line, station_name) |> 
  nrow()
distinct_A_stations
```

Of the stations that serve the A train, how many are ADA compliant?
```{r}
distinct_A_stations_ADA =
  subway_data_reformatted |> 
    filter(route_name == "A", ada == "TRUE") |>
   distinct(line, station_name) |> 
  nrow()
distinct_A_stations_ADA
```
17 Stations are ADA compliant.