---
title: "p8105_hw2_yc4959"
author: "Yimeng Cao"
date: "2026-09-30"
output: github_document
---

This is p8105 homework 2

```{r}
library(tidyverse)
```

## Problem 1

Read and clean data, select retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable.

So I need to remove the exit_only column.
```{r}
subway_data = 
  read_csv("./NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>% 
  janitor::clean_names() %>% 
  select(line:vending, - exit_only, ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

This subway dataset contains line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance.
I cleaned the variable names, selected the needed variables, removed the exit_only variable, and converted entry from character to logical.

The cleaned dataset contains `r nrow(subway_data)` rows and `r ncol(subway_data)` columns.

I don't think this is “tidy data” because the data for routes 1 through 11 is spread across different columns.


How many distinct stations are there? 
```{r}
subway_data |> 
  distinct(line, station_name) |> 
  nrow()
```
465 distinct stations are there.

How many stations are ADA compliant?
```{r}
subway_data |> 
  distinct(line, station_name, ada) |> 
  count(ada)

```
84 stations are ADA compliant.

What proportion of station entrances / exits without vending allow entrance?

Use no vending entry divided by total no vending entry
```{r}
no_vending_entry = 
  subway_data |> 
   filter(vending == "NO", entry == TRUE) |> 
  nrow()

no_vending_total =
   subway_data |> 
  filter(vending == "NO") |> 
  nrow()

proportion = no_vending_entry / no_vending_total 
proportion

```
The proportion of station entrances / exits without vending allow entrance is 37.7%.



Reformat data so that route number and route name are distinct variables.

When reshaping the data, I found that the route variables had different data types. Route1 to Route7 were stored as character variables, while Route8 to Route11 were stored as numeric variables.
So i use mutate to change the data type.
```{r}
subway_data_reformatted =
  subway_data |> 
  mutate(
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
  ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  )
```

How many distinct stations serve the A train?
```{r}
distinct_A_stations =
  subway_data_reformatted |> 
    filter(route_name == "A") |>
   distinct(line, station_name) |> 
  nrow()
distinct_A_stations
```
60 distinct stations serve the A train.

Of the stations that serve the A train, how many are ADA compliant?
```{r}
distinct_A_stations_ADA =
  subway_data_reformatted |> 
    filter(route_name == "A", ada == "TRUE") |>
   distinct(line, station_name) |> 
  nrow()
distinct_A_stations_ADA
```
17 Stations are ADA compliant.


## Problem 2
Read and clean the Mr. Trash Wheel sheet

specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel

I checked xlsx file to find the range.
```{r}
trash_df =
   readxl::read_excel(
     "./202509 Trash Wheel Collection Data.xlsx",
     sheet = "Mr. Trash Wheel",
    range = "A2:N709"
   )
```


Clena name and omit rows that do not include dumpster-specific data.
```{r}
trash_df =
  trash_df |> 
  janitor::clean_names() |> 
  drop_na(dumpster)
```

Round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
```{r}
trash_df =
  trash_df |> 
  mutate(
    sports_balls = as.integer(round(sports_balls))
  )
```


Use a similar process to import, clean, and organize the data for Professor Trash Wheel
I checked the xlsx file find the range.
```{r}
professor_df =
  readxl::read_excel(
    "./202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M134") |> 
    janitor::clean_names() |> 
    drop_na(dumpster)
```

Use a similar process to import, clean, and organize the data for Professor Trash Wheel
```{r}
gwynnda_df =
  readxl::read_excel(
    "./202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L351") |> 
    janitor::clean_names() |> 
    drop_na(dumpster)
```

To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.
First i add Mr. Trash Wheel variable
```{r}
trash_df =
  trash_df |> 
  mutate(trash_wheel = "Mr. Trash Wheel")
```

Then i add Professor Trash Wheel variable
```{r}
professor_df =
  professor_df |> 
  mutate(trash_wheel = "Professor Trash Wheel")
```

Last i add Gwynnda variable
```{r}
gwynnda_df =
  gwynnda_df |> 
  mutate(trash_wheel = "Gwynnda")
```

I found that the year variables had different data types, so I converted them to character variables
```{r}
trash_df =
  trash_df |> 
  mutate(year = as.character(year))

professor_df =
  professor_df |> 
  mutate(year = as.character(year))

gwynnda_df =
  gwynnda_df |> 
  mutate(year = as.character(year))
```

Combining
```{r}
trash_wheel_combine =
  bind_rows(
    trash_df,
    professor_df,
    gwynnda_df
  )
view(trash_wheel_combine)
```

Professor Trash Wheel total weight
```{r}
professor_weight =
  trash_wheel_combine |> 
  filter(trash_wheel == "Professor Trash Wheel") |> 
  pull(weight_tons) |> 
  sum(na.rm = TRUE)

professor_weight
```

The total number of cigarette butts collected by Gwynnda in June 2022 
```{r}
gwynnda_cigarette =
  trash_wheel_combine |> 
  filter(
    trash_wheel == "Gwynnda",
    month == "June",
    year == "2022"
  ) |> 
  pull(cigarette_butts) |> 
  sum(na.rm = TRUE)
gwynnda_cigarette
```


The combined Trash Wheel dataset contains `r nrow(trash_wheel_combine)`.
This dataset include key variables like dumpster, month, year, weight, plastic bottles, cigarette butts, and trash wheel. 
The total weight of trash collected by Professor Trash Wheel was `r professor_weight` tons.
The total number of cigarette butts collected by Gwynnda in June 2022 was `r gwynnda_cigarette`.



## Problem 3

Import, clean, and tidy the dataset of baseline demographics. 

I used skip = 1 because the first row contains variable descriptions, and the second row contains the variable names.
I also treated ".", "NA", and blank values as missing values during import.
```{r}
mci_baseline =
  read_csv("./MCI_baseline.csv",
  skip = 1, 
  na = c(".", "NA", "")) |> 
  janitor::clean_names()


```
Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric).
```{r}
mci_baseline =
  mci_baseline |> 
  mutate(
    sex = case_match(
      sex,
      0 ~ "Female",
      1 ~ "Male"
    ),sex = as.factor(sex),
    apoe4 = case_match(
      apoe4,
      0 ~ "Non-carrier",
      1 ~ "Carrier"
    ),
    apoe4 = as.factor(apoe4)
  )
view(mci_baseline)
```

Remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). 
```{r}
mci_baseline =
  mci_baseline |> 
  filter(is.na(age_at_onset) | age_at_onset > current_age)
```
How many participants were recruited?
```{r}
nrow(mci_baseline)
```

How many develop MCI?
```{r}
mci_develop =
  mci_baseline |> 
  filter(!is.na(age_at_onset)) |> 
  nrow()
mci_develop
```

What is the average baseline age?
```{r}
average_age =
  mci_baseline |> 
  pull(current_age) |> 
  mean()
average_age
```

What proportion of women in the study are APOE4 carriers?
1. female APOE4 carriers
```{r}
female_APOE4 =
  mci_baseline |> 
  filter(
    sex == "Female",
    apoe4 == "Carrier"
  ) |> 
  nrow()
female_APOE4
```
2. In total how many female
```{r}
female_total =
  mci_baseline |> 
  filter(sex == "Female") |> 
  nrow()
female_total
```

Proportion of women in the study are APOE4 carriers
```{r}
female_APOE4_proportion =
  female_APOE4 / female_total

female_APOE4_proportion
```

There were `r nrow(mci_baseline)` participants recruited in the study. 
Among these participants, `r mci_develop` developed MCI. 
The average baseline age was `r round(average_age, 2)` years. 
The proportion of women in the study who were APOE4 carriers was `r (female_APOE4_proportion * 100)`%.
The dataset contains information on participant ID, age, sex, education, APOE4 carrier status, and age at MCI onset.


Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.

I cleaned the variable names and treated missing values as NA.
```{r}
mci_amyloid =
  read_csv(
    "./mci_amyloid.csv",
    skip = 1,
    na = c(".", "NA", "")
  ) |> 
  janitor::clean_names()

```
The dataset contains amyloid measurements at baseline and at 2, 4, 6, and 8 years. I used pivot_longer() to make the dataset tidy.
```{r}
mci_amyloid =
  mci_amyloid |> 
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    values_to = "amyloid"
  )
view(mci_amyloid)
```

Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings.
Rename amyloid study_id to id so that the two datasets use the same variable name
```{r}
mci_amyloid =
  mci_amyloid |> 
  rename(id = study_id)
```

1. Check who appear in the baseline dataset but not in the amyloid dataset.
```{r}
baseline_only =
  anti_join(
    mci_baseline,
    mci_amyloid,
    by = "id"
  )
baseline_only
```
Check who appear in the amyloid dataset but not in the baseline dataset.
```{r}
amyloid_only =
  anti_join(
    mci_amyloid,
    mci_baseline,
    by = "id"
  )
amyloid_only
```

Because the amyloid data is longitudinal, so i used distinct ID to count the number of participants.
```{r}
amyloid_only |>
  distinct(id) |>
  nrow()
```

Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained
```{r}
mci_complete =
  inner_join(
    mci_baseline,
    mci_amyloid,
    by = "id"
  )
view(mci_complete)
```

Save the csv!!!!
```{r}
write_csv(
  mci_complete,
  "./mci_complete.csv"
)
```

FINISH!